A geometric framework for capturing high-resolution insights into conceptual knowledge and learning in classroom-like settings

### 2023-01-29 paper/main.tex
initial edits to results and experiment methods

Here we develop a framework for modeling how knowledge is acquired during
learning. The central idea is to use text embedding models to define the
coordinate systems of two maps: (a) a \textit{knowledge map} that describes the
extent to which each concept is currently known and (b) a \textit{learning map}
that describes changes in knowledge over time. Each location on these maps
represents a single concept, and the maps' geometries are defined such that
related concepts are located nearby in space. We use this framework to analyze
and interpret behavioral data collected from an experiment that had
participants watch and answer multiple choice questions about a series of
recorded course lectures.

Our primary research goal is to advance our understanding of what it means to
acquire deep real-world conceptual knowledge. Traditional laboratory approaches
to studying learning and memory (e.g., list learning studies) often draw little
distinction between memorization and understanding. Instead, these studies
typically focus on whether information is effectively encoded or retrieved,
rather than whether the information is \textit{understood}. Approaches to
studying conceptual learning, such as category learning experiments, can start
to investigate the distinction between memorization and understanding, often by
training participants to distinguish arbitrary or random features in otherwise
meaningless categorized stimuli. However the objective of real-world training,
or learning from life experiences more generally, is often to develop new
knowledge that may be applied in \textit{useful} ways in the future. In this
sense, the gap between modern learning theories and modern pedagogical
approaches and classroom learning strategies is enormous: most of our theories
about \textit{how} people learn are inspired by experimental paradigms and
models that have only peripheral relevance to the kinds of learning that
students and teachers actually seek \citep{Macl05}. To help bridge this gap, our study uses
course materials from real online courses to inform, fit, and test models of
real-world conceptual learning. We also provide a ``proof of concept''
demonstration of how our models might be used to construct ``maps'' of what
students know, and how their knowledge changes with training. In addition to
helping to visualize knowledge (and changes in knowledge), we hope that such
maps might lead to real-world tools for improving how we educate.

@@

    \caption{\textbf{Experimental paradigm.} Participants alternate between
    answering 13-question multiple choice quizzes and watching two Khan academy
    videos. Each quiz contains a mix of 5 questions about lecture 1, 5
    questions about lecture 2, and 3 general physics knowledge questions. The
    specific questions reflected on each quiz, and the orders of each quiz's
    questions, were randomized across participants.}

@@

We asked our participants to answer questions from several multiple choice
quizzes and watch two lecture videos from the \textit{Khan Academy} platform
(Fig.~\ref{fig:experiment}). One lecture video, entitled \textit{Four
Fundamental Forces}, was about the four fundamental forces in physics: gravity,
strong and weak interactions, and electromagnetism. The second lecture video,
entitled \textit{Birth of Stars}, provides an overview of our current
understanding of how stars form. We selected both lessons to be (a) accessible
to a broad audience, e.g., by minimizing prerequisite knowledge, (b) largely
independent of each other, e.g., so that the two videos focused on different
material and did not depend on each other, and (c) related to each other, e.g.,
so that both videos contained at least \textit{some} similar or overlapping
content. The two videos we selected are introductory (i.e., minimizing specific
prerequisite knowledge), and are about different primary concepts, but they
both also touch on ``physics'' and ``astronomy'' themes. We also wrote a set of
multiple choice quiz questions that would enable us to test participants'
knowledge about each individual video and about related content not
specifically presented in either video (Tab.~\questions). Participants answered
questions randomly drawn from each content area (lecture 1, lecture 2, and
general physics knowledge) across each of three quizzes. Quiz 1 was intended to
assessed participants' knowledge before training; quiz 2 assessed knowledge
after watching the \textit{Four Fundamental Forces} video (i.e., lecture 1);
and quiz 3 assessed knowledge after watching the \textit{Birth of Stars} video
(i.e., lecture 2).

We trained a text embedding model using sliding windows of text from the two
videos' transcripts (see \textit{Constructing text embeddings of multiple
videos and questions}). We also used the same model (i.e., trained on the
videos' transcripts) to embed the text of each question in our pool. This
yielded, for each second of each video, and for each question, a single
15-dimensional topic vector---i.e., a coordinate in a text embedding space
(Fig.~\ref{fig:sliding-windows}). Intuitively, each dimension of the embedding
space corresponds to a ``theme'' or ``topic'' reflected in some part(s) of the
videos (Tab.~\topics), and the coordinates in embedding space denote the blend
of themes reflected by a particular excerpt of text (e.g., from part of a
video's transcript, from a question, etc.).

\begin{figure}[tp]
    \centering
    \includegraphics[width=0.6\textwidth]{figs/active-topics}

\caption{\textbf{Lecture and question topic overlap.} The bar plots display the
variability in topic weights across lecture timepoints (top panels) and
questions (bottom panels); colors denote topics. The top-weighted words from
the most ``expressive'' (i.e., variable across observations) topics from each
lecture are displayed in the upper right (orange: topic 2; yellow-green: topic
4). The top-weighted words from the full set of topics may be found in
Table~\topics.}

@@

We developed a simple formula (Eqn.~\ref{eqn:prop}) for using a participant's
responses to a small set of multiple choice questions to estimate how much the
participant ``knows'' about the concept reflected by any arbitrary coordinate,
$x$, in text embedding space (e.g., the content reflected by any moment in a
lecture they had watched; see \textit{Estimating dynamic knowledge traces}).
Essentially, the estimated knowledge at the coordinate is given by the weighted
average proportion of quiz questions the participant answered correctly, where
the weights reflect how much each question is ``about'' the content at $x$.
When we apply this approach to estimate the participant's knowledge about the
content presented in each moment of each lecture, we can obtain a detailed
timecourse describing how much ``knowledge'' the participant has about any part
of the lecture. As shown in Figure~\ref{fig:knowledge-timeseries}, we can also
apply this approach separately for the questions from each quiz the
participants took throughout the experiment. From just 13 questions per quiz,
we obtain a high-resolution snapshot (at the time each quiz was taken) of what
the participants knew about any moment's content, from either of the two
lectures they watched (comprising a total of 1106 samples across the two
lectures).

@@

Participants reported their undergraduate major(s) as Social Sciences (28
participants), Natural sciences (16), Professional (e.g., pre-med or pre-law; 8
participants), Mathematics and engineering (7 participants), Humanities (4
participants), or Undecided (3 participants). Note that some participants
selected multiple categories for their undergraduate major. We also asked
participants about the courses they had taken. In total, 46 participants
reported having taken at least one Khan academy course in the past or being
familiar with the Khan academy, and 4 reported not having taken any Khan
academy courses. Of the participants who reported having watched at least one
Khan academy course, 1 participant declined to report the number of courses
they had watched; 7 participants reported having watched 1--2 courses; 11
reported having watched 3--5 courses; 8 reported having watched 5--10 courses;
and 19 reported having watched 10 or more courses. We also asked participants
about the specific courses they had watched, categorized under different
subject areas. In the ``Mathematics'' area participants reported having watched
videos on AP Calculus AB (21 participants), Precalculus (17 participants),
Algebra 2 (14 participants), AP Calculus BC (12 participants), Trigonometry (11
participants), Algebra 1 (10 participants), Geometry (8 participants),
Pre-algebra (7 participants), Multivariable Calculus (5 participants),
Differential Equations (5 participants), Statistics and Probability (4
participants), AP Statistics (2 participants), Linear Algebra (2 participants),
Early Math (1 participant), Arithmetic (1 participant), and other videos not
listed in our survey (6 participants). In the ``Science and engineering'' area
participants reported having watched videos on Chemistry, AP Chemistry, or
Organic Chemistry (21 participants); Physics, AP Physics I, or AP Physics II (15
participants); Biology, AP Biology; or High school Biology (15 participants);
Health and Medicine (1 participant); or other videos not listed in our survey
(20 participants). We also asked participants if they had specifically seen the
videos used in our experiment. When we asked about the \textit{Four Fundamental
Forces} video, 45 participants reported not having watched it before, 1
participant reported that they were not sure if they had watched it before, and
4 participants declined to respond. When we asked about the \textit{Birth of
Stars} video, 46 participants reported not having watched it before and 4
participants declined to respond. When we asked participants about non-Khan
academy online courses, they reported having watched or taken courses on
Mathematics (15 participants), Science and engineering (11 participants), Test
preparation (9 participants), Economics and finance (3 participants), Arts and
humanities (2 participants), Computing (2 participants), and other categories
not listed in our survey (18 participants). Finally, we asked participants
about in-person courses they had taken in different subject areas. They
reported taking courses in Mathematics (39 participants), Science and
engineering (38 participants), Arts and humanities (35 participants), Test
preparation (27 participants), Economics and finance (26 participants),
Computing (15 participants), College and careers (7 participants), or other
courses not listed in our survey (6 participants).

@@

We hand-selected two course videos from the Khan Academy platform: \textit{Four
Fundamental Forces} (an introduction to gravity, electromagnetism, the weak
nuclear force, and the strong nuclear force; duration: 10 minutes and 29
seconds) and \textit{Birth of Stars} (an introduction to how stars are formed;
duration: 7 minutes and 57 seconds). We hand-wrote 39 multiple choice
questions: 15 about the conceptual content of \textit{Four Fundamental Forces},
another 15 about the conceptual content of \textit{Birth of Stars}, and 9 other
questions that tested for general conceptual knowledge about basic physics
(covering material that was not presented in either video). The full set of
questions may be found in Table~\questions.

Participants began the main experiment by answering a battery of 13 randomly
selected questions (chosen from the full set of 39). Then they watched the
\textit{The Four Fundamental Forces} video. Next, they answered a second set of
13 questions (chosen at random from the remaining 26 questions). Fourth,
participants watch the \textit{Birth of Stars} video, and finally they answered
the remaining 13 questions. Our experimental procedure is diagramed in
Figure~\ref{fig:experiment}. We used the experiment to develop and test our
computational framework for estimating knowledge and learning.

@@

\subsubsection*{Constructing text embeddings of multiple videos and questions}

>>>>

Here we develop a framework for modeling how knowledge is acquired during
learning. The central idea is to use text embedding models to define the
coordinate systems of two maps: (a) a \textit{knowledge map} that describes the
extent to which each concept is currently known and (b) a \textit{learning map}
that describes changes in knowledge over time. Each location on these maps
represents a single concept, and the maps' geometries are defined such that
related concepts are located nearby in space. We use this framework to analyze
and interpret behavioral data collected from an experiment that had
participants watch and answer multiple-choice questions about a series of
recorded course lectures.

Our primary research goal is to advance our understanding of what it means to
acquire deep real-world conceptual knowledge. Traditional laboratory approaches
to studying learning and memory (e.g., list learning studies) often draw little
distinction between memorization and understanding. Instead, these studies
typically focus on whether information is effectively encoded or retrieved,
rather than whether the information is \textit{understood}. Approaches to
studying conceptual learning, such as category learning experiments, can start
to investigate the distinction between memorization and understanding, often by
training participants to distinguish arbitrary or random features in otherwise
meaningless categorized stimuli. However the objective of real-world training,
or learning from life experiences more generally, is often to develop new
knowledge that may be applied in \textit{useful} ways in the future. In this
sense, the gap between modern learning theories and modern pedagogical
approaches and classroom learning strategies is enormous: most of our theories
about \textit{how} people learn are inspired by experimental paradigms and
models that have only peripheral relevance to the kinds of learning that
students and teachers actually seek \citep{Macl05, HallGree08}. To help bridge this gap, our study uses
course materials from real online courses to inform, fit, and test models of
real-world conceptual learning. We also provide a ``proof of concept''
demonstration of how our models might be used to construct ``maps'' of what
students know, and how their knowledge changes with training. In addition to
helping to visualize knowledge (and changes in knowledge), we hope that such
maps might lead to real-world tools for improving how we educate.

@@

    \caption{\textbf{Experimental paradigm.} Participants alternate between
    answering 13-question multiple-choice quizzes and watching two Khan Academy
    lectures. Each quiz contains a mix of 5 questions about lecture 1, 5
    questions about lecture 2, and 3 questions about general physics knowledge. The
    specific questions reflected on each quiz, and the orders of each quiz's
    questions, were randomized across participants.}

@@

We asked participants in our study to answer questions from several multiple-choice
quizzes and watch two lecture videos from the Khan Academy platform
(Fig.~\ref{fig:experiment}). The first lecture video, entitled \textit{Four
Fundamental Forces}, discussed the four fundamental forces in physics: gravity,
strong and weak interactions, and electromagnetism. The second,
entitled \textit{Birth of Stars}, provided an overview of our current
understanding of how stars form. We selected these lessons to be (a) accessible
to a broad audience, i.e., requiring minimal prerequisite knowledge to understand;
(b) conceptually related to each other, i.e., covering at least \textit{some}
similar or overlapping content; and (c) largely independent of each other, i.e., focused
on sufficiently different material that understanding one did not require having seen
the other. Thus, the two videos we selected are introductory, high-level lectures that both 
belong to Khan Academy's ``Cosmology and Astronomy" course domain, but come from
different lecture series (``Scale of the Universe" and ``Stars, Black Holes, and Galaxies" 
for the first and second lectures, respectively).

We also wrote a set of
multiple-choice quiz questions that would enable us to test participants'
knowledge about each individual lecture, as well as related content not
specifically presented in either video ( see Tab.~\questions). Participants answered
questions randomly drawn from each content area (lecture 1, lecture 2, and
general physics knowledge) across each of three quizzes. Quiz 1 was intended to
assessed participants' ``baseline" knowledge before training, quiz 2 assessed knowledge
after watching the \textit{Four Fundamental Forces} video (i.e., lecture 1),
and quiz 3 assessed knowledge after watching the \textit{Birth of Stars} video
(i.e., lecture 2).

We trained a topic model \citep{BleiEtal03} using sliding windows of text from the two
lectures' transcripts (see \hyperref[subsec:topic-modeling]{\textit{Constructing text embeddings of multiple
lectures and questions}}). We also used the same model (i.e., trained on the
videos' transcripts) to embed the text of each question in our pool. This
yielded, for each second of each video, and for each question, a single
15-dimensional topic vector---i.e., a coordinate in a text embedding space
(Fig.~\ref{fig:sliding-windows}). Intuitively, each dimension of the embedding
space corresponds to a ``theme'' or ``topic'' reflected in some part(s) of the
videos (Tab.~\topics), and the coordinates in embedding space denote the blend
of themes reflected by a particular excerpt of text (e.g., from part of a
video's transcript, from a question, etc.).

\begin{figure}[tp]
    \centering
    \includegraphics[width=0.7\textwidth]{figs/active-topics}

\caption{\textbf{Lecture and question topic overlap.} The bar plots display the
variability in topic weights across lecture timepoints (top panels) and
questions (bottom panels); colors denote topics. The top-weighted words from
the most ``expressive'' (i.e., variable across observations) topic from each
lecture are displayed in the upper right (orange: topic 2; yellow-green: topic
4). The top-weighted words from the full set of topics may be found in
Table~\topics.}

@@

We developed a simple formula (Eqn.~\ref{eqn:prop}) for using a participant's
responses to a small set of multiple-choice questions to estimate how much the
participant ``knows'' about the concept reflected by any arbitrary coordinate,
$x$, in text embedding space (e.g., the content reflected by any moment in a
lecture they had watched; see \textit{Estimating dynamic knowledge traces}).
Essentially, the estimated knowledge at the coordinate is given by the weighted
average proportion of quiz questions the participant answered correctly, where
the weights reflect how much each question is ``about'' the content at $x$.
When we apply this approach to estimate the participant's knowledge about the
content presented in each moment of each lecture, we can obtain a detailed
timecourse describing how much ``knowledge'' the participant has about any part
of the lecture. As shown in Figure~\ref{fig:knowledge-timeseries}, we can also
apply this approach separately for the questions from each quiz the
participants took throughout the experiment. From just 13 questions per quiz,
we obtain a high-resolution snapshot (at the time each quiz was taken) of what
the participants knew about any moment's content, from either of the two
lectures they watched (comprising a total of 1106 samples across the two
lectures).

@@

Participants reported their undergraduate major(s) as ``social sciences" (28
participants), ``natural sciences" (16 participants), ``professional" (e.g., pre-med or pre-law; 8
participants), ``mathematics and engineering" (7 participants), ``humanities" (4
participants), or ``undecided" (3 participants). Note that some participants
selected multiple categories for their undergraduate major. We also asked
participants about the courses they had taken. In total, 45 participants
reported having taken at least one Khan Academy course in the past, 
and 5 reported not having taken any Khan
Academy courses. Of those who reported having watched at least one
Khan Academy course, 7 participants reported having watched 1--2 courses, 11
reported having watched 3--5 courses, 8 reported having watched 5--10 courses,
and 19 reported having watched 10 or more courses. We also asked participants
about the specific courses they had watched, categorized under different
subject areas. In the ``Mathematics'' area, participants reported having watched
videos on AP Calculus AB (21 participants), Precalculus (17 participants),
Algebra 2 (14 participants), AP Calculus BC (12 participants), Trigonometry (11
participants), Algebra 1 (10 participants), Geometry (8 participants),
Pre-algebra (7 participants), Multivariable Calculus (5 participants),
Differential Equations (5 participants), Statistics and Probability (4
participants), AP Statistics (2 participants), Linear Algebra (2 participants),
Early Math (1 participant), Arithmetic (1 participant), and other videos not
listed in our survey (5 participants). In the ``Science and engineering'' area,
participants reported having watched videos on Chemistry, AP Chemistry, or
Organic Chemistry (21 participants); Physics, AP Physics I, or AP Physics II (15
participants); Biology, AP Biology; or High school Biology (15 participants);
Health and Medicine (1 participant); or other videos not listed in our survey
(19 participants). We also asked participants whether they had specifically seen the
videos used in our experiment. Of the 45 participants who reported having having
taken at least one Khan Academy course in the past, 44 participants reported that they had not watched the \textit{Four Fundamental Forces} video, and 1 participant reported that they were not sure whether they had watched it. All participants reported that they had not watched the \textit{Birth of Stars} video.
When we asked participants about non-Khan
Academy online courses, they reported having watched or taken courses on
Mathematics (15 participants), Science and engineering (11 participants), Test
preparation (9 participants), Economics and finance (3 participants), Arts and
humanities (2 participants), Computing (2 participants), and other categories
not listed in our survey (18 participants). Finally, we asked participants
about in-person courses they had taken in different subject areas. They
reported taking courses in Mathematics (39 participants), Science and
engineering (38 participants), Arts and humanities (35 participants), Test
preparation (27 participants), Economics and finance (26 participants),
Computing (15 participants), College and careers (7 participants), or other
courses not listed in our survey (6 participants).

@@

We hand-selected two course videos from the Khan Academy platform: \textit{Four
Fundamental Forces} (an introduction to gravity, electromagnetism, the weak
nuclear force, and the strong nuclear force; duration: 10 minutes and 29
seconds) and \textit{Birth of Stars} (an introduction to how stars are formed;
duration: 7 minutes and 57 seconds). We hand-wrote 39 multiple-choice
questions: 15 about the conceptual content of \textit{Four Fundamental Forces},
15 about the conceptual content of \textit{Birth of Stars}, and 9
questions that tested for general conceptual knowledge about basic physics
(covering material that was not presented in either video). The full set of
questions and answer options may be found in Table~\questions.

%Over the course of the experiment, participants completed three multiple-choice quizzes

Participants began the main experiment by answering a battery of 13 randomly
selected questions (chosen from the full set of 39). Then they watched the
\textit{The Four Fundamental Forces} lecture video. Next, they answered a second set of
13 questions (chosen at random from the remaining 26 questions). Fourth,
participants watch the \textit{Birth of Stars} video, and finally they answered
the remaining 13 questions. Our experimental procedure is diagramed in
Figure~\ref{fig:experiment}. We used the experiment to develop and test our
computational framework for estimating knowledge and learning.

@@

\subsubsection*{Constructing text embeddings of multiple lectures and questions}\label{subsec:topic-modeling}
